Using structural knowledge to correct RNA pyrosequencing errors in non-coding RNAs

### 2013-05-22 JCB/methods_JCB.tex
Finished outside

The \emph{Inside} function $\Z{\Struct}{m}{x,y}$ is simply the partition function, i.e. the sum of Boltzmann factors over all sequences for a substructure $\Struct$ (implicitly attached to an interval $[i,j]$ of the sequence), featuring $m$ mutations/errors compared to $s$, and having flanking nucleotides $x$ and $y$. 
Such terms can be computed by recurrence, using the following equation for the initial case:
\begin{equation}
	\forall x,y\in B\times \B, m\in [0,M]:\, \Z{\varepsilon}{m}{x,y}=\left\{
	\begin{array}{ll}
		1 &\text{ If } m = 0\\
		0 &\text{ Otherwise.}
	\end{array}\right.
\label{eq:Z_in}
\end{equation}
In other words, either there is no sequence at distance $m>0$ of the empty sequence, or the only allowed sequence is the empty sequence ($m=0$), having energy $0$. Since the contributions only depend on base pairs, they do not appear in the initial conditions.

@@

\subsubsection{Outside computation.}	
\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{FigDPOutsideWrapper}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires a precomputation of the inside contributions.}
\end{figure}

The \emph{Outside} function, $\Y{\Struct}{m}{x,y}$ is the partition function considering only the 
contributions of subsequences excluding a given structure/interval $\Struct$, occupying the open interval $]i,j[$ in the sequence, at Hamming distance exactly $m$ to the initial sequence $s$, and assuming that nucleotides $x$ and $y$ were previously chosen for $i+1$ and $j-1$, the outermost portions of the excluded structure.
\TODOYann{Reformat outside Figure as in Incarnation.}
The associated terms $\Y{\Struct}{m}{x,y}$ can then be computed recursively, considering as the base case any prefix $\Struct$ of the complete structure:
\begin{equation}
	\forall x,y\in B\times \B, \forall \Struct'\text{ s.t. }\Struct=\Struct'.\Struct'', m\in [0,M]:\, \Y{\Struct^\ast}{m}{x,y}=\Z{\Struct''}{m}{y,X}
\label{eq:Y_in}
\end{equation}
where $X\in\B$ can be any nucleotide, and will not affect further computations.

The main recurrence below extends the excluded structure leftward, and considers four cases depending on its base-pairing status and directionality:
\begin{itemize}
\item {\bf Case 1: Unpaired position.}
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'\in \B,\\ \Kron_{a,a'}\le m}}
    \Y{\ub\Struct}{m- \Kron_{a,a'}}{a',y}
\label{eq:Y_rec_A}
\end{equation}
\item {\bf Case 2: Stacking base-pair.}
\begin{equation}
	\Y{\Struct}{m}{x,y} = 
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 e^{\frac{-\EBP{i,j}{xy}{a'b'}}{RT}}\cdot
    \Y{\op\Struct\cp}{m- \Kron_{ab,a'b'}}{a',b'} 
\label{eq:Y_rec}
\end{equation}
\item {\bf Case 3: Next position paired rightward.}
\begin{equation}
	\Y{i,j}{m}{a,b} = \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
  		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{i-1,k+1}{m- \Kron_{a'b',s_is_k} - m'}{a',b'}
     \cdot\Z{j,k-1}{m'}{b,b'} 
\label{eq:Y_rec}
\end{equation}
\item {\bf Case 4: Next position paired leftward.}
\begin{equation}
	\Y{i,j}{m}{a,b} = 
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_ks_i}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_ks_i}}
   	 e^{\frac{-E_{(k,i),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{k-1,j}{m- \Kron_{a'b',s_ks_i} - m'}{a',b}
     \cdot\Z{k+1,i-1}{m'}{a',b'}
\label{eq:Y_rec}
\end{equation}
\end{itemize}
\TODOYann{Rephrase in term of tree, as in Incarnation??}
The five cases can be broked down as follows.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we decrease the lower interval bound by $1$ (i.e. $i-1$), and consider all possible
nucleotides $a'$ at position $i$, correcting the number of mutants
in function of $\Kron_{a',s_i}$.
\item[$S_{i}=j$ and $S_{i+1}=j-1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i+1$ is
paired with $j-11$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{outside} value for the interval $[i-1,j+1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_{i}=k \geq j$:]If nucleotide $i$ is paired with position $k\geq j$, 
and is not stacked inside, the 
only term contributing directly to the energy is the isostericity of base pair $(i,k)$. 
Therefore, we consider the outside interval $[i-1,k+1]$, multiplying it by the \emph{inside}
value of the newly included interval (i.e. $[j,k-1]$), for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $k$ respectively.
\item[$-1<S_{i}<i$:]As above but if position $i$ is paired with a lower value.
\item[Else:] Any other derivation of the SCFG does not correspond to the 
secondary structure $S$, and we return $0$.

\end{description}

\subsubsection{Combining Inside and Outside Values into Point-Wise Mutations Probabilities.}
By construction, the partition function over all sequences at exactly $m$ mutations of a reference sequence $s$ can 
be either described from the \emph{inside} contribution $\Z{0,n-1}{m}{X,X}$ of the whole sequence,
$\forall X\in B$, or from \emph{outside} terms as:
$$
	\Z{0,n-1}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a},\forall k \text{	unpaired.}
$$

We are now left to compute the probability that a given position is a given nucleotide.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations, one defines
\begin{equation}
	\mathbb{P}(s_i = x\;|\; M) := \frac{\mathcal{W}^M_{\substack{i, [x]}}}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\label{eq:normalize}
\end{equation}
where $\mathcal{W}{\substack{\ast\\ \ast}}$ is defined by:
\begin{equation}
\resizebox{\textwidth}{!}{%
$ \mathcal{W}^M_{\substack{i, [x]}} =  \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\cdot\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\cdot\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
    \displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
			\cdot\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\cdot\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.
$}\label{eq:combine}
\end{equation}

@@

\subsection{Complexity Considerations}
Equations~\eqref{eq:Z_rec} and~\eqref{eq:Y_rec} can be computed using dynamic programming. Namely, the $\mathcal{Z}^{*}_{*}$ and $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and interval lengths, memorizing the results as they become available to ensure constant-time access during later stages of the computation. Furthermore, energy terms $E(\cdot)$ can be accessed in constant time thanks to a simple precomputation (not described)  of the isostericity contributions in $\Theta(n\cdot|\Omega|)$. Computing any given term therefore requires $\Theta(m)$ operations.

>>>>

The \emph{Inside} function $\Z{\Struct}{m}{x,y}$ is simply the partition function, i.e. the sum of Boltzmann factors over all sequences for a substructure $\Struct$ (implicitly attached to an interval $[i,j]$ of the sequence), featuring $m$ mutations/errors compared to $s$, and having flanking nucleotides $x$ and $y$. 
Such terms can be computed recursively, using the following equation for the initial case:
\begin{equation}
	\forall x,y\in B\times \B, m\in [0,M]:\, \Z{\varepsilon}{m}{x,y}=\left\{
	\begin{array}{ll}
		1 &\text{ If } m = 0\\
		0 &\text{ Otherwise.}
	\end{array}\right.
\label{eq:Z_in}
\end{equation}
In other words, either there is no sequence at distance $m>0$ of the empty sequence, or the only allowed sequence is the empty sequence ($m=0$), having energy $0$. Since the contributions only depend on base pairs, they do not appear in the initial conditions.

@@

\subsubsection{Outside computation.}

The \emph{Outside} function, $\Y{\Struct}{m}{x,y}$ is the partition function considering only the 
contributions of subsequences excluding a given structure/interval $\Struct$, occupying the open interval $]i,j[$ in the sequence, at Hamming distance exactly $m$ to the initial sequence $s$, and assuming that nucleotides $x$ and $y$ were previously chosen for $i+1$ and $j-1$, the outermost portions of the excluded structure.
\TODOYann{Reformat outside Figure as in Incarnation.}
The associated terms $\Y{\Struct}{m}{x,y}$ can then be computed recursively, initially considering the case of any prefix $\Struct'$ of the complete structure:
\begin{equation}
	\forall x,y\in \B\times \B, \forall \Struct'\text{ s.t. }\Struct=\Struct'.\Struct'', m\in [0,M]:\, \Y{\Struct'}{m}{x,y}=\Z{\Struct''}{m}{y,z}
\label{eq:Y_in}
\end{equation}
where $z\in\B$ can be any nucleotide, and provably does not affect further computations. 
In other words, the sequences explored by an outside computation, excluding a prefix of $\Struct$, are exactly the sequences generated on the corresponded suffix. This set of sequence is also the inside term on the suffix structure.
It is also worth pointing out that $\Y{\Struct}{m}{x,y}= \Z{\varepsilon}{m}{y,X}=\mathbbm{1}_{m=0}$.

\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{FigDPOutsideWrapper}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires the precomputation and storage of the inside contributions.}
\end{figure}
In general, the main recurrence below works by extending the excluded structure $\Struct$ (covering the $]i,j[$ interval) in the leftward direction, and considers four cases depending on the base-pairing status and directionality of the next considered position:
\begin{itemize}
\item {\bf Case 1: Unpaired position.} When the innermost leftward base is unpaired, any base $a\in\B$ may be chosen for this position, and the rest of the sequence is generated recursively. The number of remaining mutations may be decreased by the choice of $a$, leading to the following recurrence:
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'\in \B,\\ \Kron_{a,a'}\le m}}
    \Y{\ub\Struct}{m- \Kron_{a,a'}}{a',y}
\label{eq:Y_rec_A}
\end{equation}
\item {\bf Case 2: Stacking base-pair.} If both ends of the excluded structure are paired together, and are nested within another base-pair in the remaining structure, then an additional contribution for stacking has to be considered. The outside terms are then computed by simulating the choice of two base-pairing nucleotides, and by recurring on the remaining portion, as follows:
\begin{equation}
	\Y{\Struct}{m}{x,y} = 
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 e^{\frac{-\EBP{i,j}{xy}{a'b'}}{RT}}\cdot
    \Y{\op\Struct\cp}{m- \Kron_{ab,a'b'}}{a',b'} 
\label{eq:Y_rec_B}
\end{equation}
\item {\bf Case 3: Next position paired rightward, in the absence of stacking pair.} In this case, the innermost leftward position $i$ is paired to the right at some position $k$ . 
Let us assume that its partner resides outside the excluded structure $\Struct$ (This assumption is provably without loss of generality, and directly follows from the fact that $\Struct$ is well-parenthesized). 
The structure within the base-pair may be described by the expression $\op\Struct \Struct'\cp$, where $\Struct'$ may possibly be empty (except if $\Struct$ is enclosed within corresponding brackets, to avoid stacking pairs). Therefore, any sequence considered by the outside computation consists in three independent parts: two nucleotides for the paired positions, a sequence for the region excluding $\Struct''$, and a sequence for $\Struct'$. It follows that:
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
  		 e^{\frac{-\EBP{i,k}{\varnothing}{a'b'}}{RT}}
		 \cdot\Y{\op\Struct \Struct'\cp}{m- \Kron_{ab,a'b'} - m'}{a',b'}
     \cdot\Z{\Struct'}{m'}{y,b'} .
\label{eq:Y_rec_C}
\end{equation}
\item {\bf Case 4: Next position paired leftward.} If the innermost, leftward, position $i$ is paired to some position $k<i$, delimiting a substructure $\Struct'$, then any sequence considered by the outside computation consists in three parts: two nucleotides $a$ and $b$, a sequence for $\Struct'$, and a sequence for the region excluding $\op\Struct'\cp\Struct$. Consequently, one has:
\begin{equation}
	\Y{i,j}{m}{x,y} = 
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
   	 e^{\frac{-\EBP{k,i}{\varnothing}{a'b'}}{RT}}
		 \cdot\Y{\op\Struct'\cp\Struct}{m- \Kron_{ab,a'b'} - m'}{a',b}
     \cdot\Z{\Struct'}{m'}{a',b'}
\label{eq:Y_rec_D}
\end{equation}
\end{itemize}

\subsubsection{Combining Inside and Outside Computations into Point-Wise Mutations Probabilities.}

We are now left to compute the probability that a a given nucleotide $a\in\B$ is found at a given position $i$.
This quantity can also be expressed as the ratio of $\mathcal{W}^M_{\substack{i, [a]}}$, the total Boltzmann weight 
of the set of sequences featuring the nucleotide, and $\Z{\varepsilon}{\le M}{X,X}$ the total weight of sequences having at most $M$ mutations:
\begin{equation}
	\mathbb{P}(s_i = a\;|\; M) := \frac{\mathcal{W}^M_{\substack{i, [a]}}}{\sum_{m=0}^{M}\Z{\varepsilon}{m}{X,X}}\label{eq:normalize}
\end{equation}

To that purpose, we leverage the \emph{Inside-Outside} construction, and obtain the three following cases:
\begin{itemize}
\item 
where $\mathcal{W}{\substack{\ast\\ \ast}}$ is defined by:
\begin{equation}
\resizebox{\textwidth}{!}{%
$ \mathcal{W}^M_{\substack{i, [x]}} =  \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\cdot\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\cdot\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
    \displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
			\cdot\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\cdot\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.
$}\label{eq:combine}
\end{equation}
\end{itemize}

@@

By construction, the partition function over all sequences at exactly $m$ mutations of a reference sequence $s$ can 
be either described from the \emph{inside} contribution $\Z{\Struct}{m}{X,X}$ of the whole sequence,
$\forall X\in B$, or from a subset of the \emph{outside} terms, such that:
$$
	\Z{\Struct}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{\ub}{m-\Kron_{a,s[k]}}{a,a},\forall k \text{	unpaired.}
$$

\subsection{Complexity Considerations}
Equations~\eqref{eq:Z_rec_A}-\eqref{eq:Z_rec_C} and~\eqref{eq:Y_rec_A}-\eqref{eq:Y_rec_D} can be computed using dynamic programming. Namely, the $\mathcal{Z}^{*}_{*}$ and $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and interval lengths, memorizing the results as they become available to ensure constant-time access during later stages of the computation. Furthermore, energy terms $E(\cdot)$ can be accessed in constant time thanks to a simple precomputation (not described)  of the isostericity contributions in $\Theta(n\cdot|\Omega|)$. Computing any given term therefore requires $\Theta(m)$ operations.
